|

### 2017-10-07 manuscript.Rmd

Docker is a popular open source tool to create, distribute, deploy,
and run software applications using *containers*.  Containers provide a
virtual environment (see \citet{Clark2014} for an overview of common virtual environments) requiring all operating-system components an application needs to run:
code, runtime, system tools, system runtime.  Docker containers are
lightweight as they share the operating system kernel, starting instantly
using a layered filesystem which minimizes disk footprint and download time,
are built on open standards that run on all major platforms (Linux, Mac,
Windows), and provide an added layer of security by running an application
in an isolated environment \citep{what-docker}.  Familiarity with a few key
terms is helpful in understanding this paper. The term "container" refers to 
an isolated software environment on a computer. R users can think of running
a container as analogous to loading an R package; it makes a set of software
functions available.  A Docker "image" is a binary archive of that software,
analogous to an R binary package -- a given version need be downloaded only
once and then can be "run" to create a container whenever it is needed.
A "Dockerfile" is a recipe, the source-code, to create a Docker image.  Development
and contributions to the Rocker project focuses on the construction, organization
and maintenance of these Dockerfiles.

@@

@@

Importantly, Rocker containers are just as easy to run on locally on most modern laptops
by first installing the appropriate Docker distribution for Windows, MacOS, and Linux-based
operating systems. By sharing volumes with the local host, users can still manipulate
files with familiar, native tools while performing computation through a reproducible,
containerized environment \citep{Boettiger2015}. Being
able to test code in a predictable, pre-configured R environment on a
local machine and to then run the same code in an identical environment on a
remote server (e.g. for access to greater RAM, more processors, or merely
to free up the local machine from a long-running computation) is essential
for low-friction scaling of analysis.  Without such containerization,
getting code to run appropriately in a remote environment can be a major
undertaking, requiring both time and knowledge many would-be users may
not have.

@@

This sandboxing feature is also valuable in the 
remote computing context, allowing system administrators to grant
users freedom to install software which requires root privileges inside a container,
while not granting them root access on the host machine. Root access
is required to launch Docker containers, though not to access containers already
running and providing some service such as RStudio. Users logging into a container
through the RStudio\textsuperscript{\textregistered} interface do not by default have root privileges, though are able
to install R packages. Granting these users root privileges in the container still leaves
them sandboxed from the host container.  Further, Docker images can
be deployed from a machine without root privileges through the alternate container
runtime environment, Singularity\citep{singularity}.  Unlike traditional virtual machines,
these containers do not impose a large footprint of reserved resources --
a typical host can easily support 100s of containers \cite{what-docker}.

@@

Users building on the version-tagged images will by default use the MRAN snapshot mirror \citep{MRAN}
associated with the most recent date for which that image was current.  This ensures that
a Dockerfile building `FROM rocker/verse:3.4.0` will only install R package versions that were
available on CRAN on 2017-06-30, that is, the day R v3.4.1 was released.  This default
can of course be overwritten in the standard R manner, e.g. by specifying a different CRAN mirror explicitly in any 
command to `install.packages()`, or by adjusting the default CRAN mirror in `options(repo=<CRAN-MIRROR>)`
in an `.Rprofile`.  Note that the MRAN date associated with the current release (e.g. `3.4.1` at the time of
writing) will continue to advance on the Docker-hub image until the next R release.  Software
installed from `apt-get` in these images will come from the the stable Debian release (`stretch` or `jessie`)
and thus not change versions (though it will receive security patches).  Packages installed from BioConductor
using `bioclite()` utility will also install the version appropriate to the version of R found on the system
(the Bioconductor semi-annual release model avoids the need for an MRAN mirror). Users installing packages
from GitHub or other sources can request a specific git release tag or hash for a more reproducible build,
or adopt an alternative approach such as `packrat` \citep{packrat}. A more general discussion of the use and limitations of Docker for computational reproducibility can be found in \citep{Boettiger2015}. 
<!--
however, `packrat` should not be necessary
for just maintaining reproducibility with respect to CRAN and Bioconductor stacks installed on versioned images.
-->

@@

Any portable computational environment faces an inevitable tension between
the "kitchen sink problem" at one extreme, and the "discovery problem" on
the other.  A kitchen sink image seeks to accommodate too many use cases
in a single image.  Such images are inevitably very large and thus slow
or difficult to deploy, maintain and optimize.  At the other extreme, providing
too many specialized images makes it more difficult for a user to discover 
the one they need.  The Rocker project seeks to avoid both of these problems
by providing a carefully curated suite of images that an be easily extended by
individuals and communities.

@@

The Rocker project consists of a suite of images built automatically by
and hosted on the Docker Hub, <https://hub.docker.com/r/rocker>.  Source
Dockerfiles, supporting scripts and documentation are hosted on GitHub
under the organization `rocker-org`, <https://github.com/rocker-org>.
The issue tracker and pull requests are used for community input,
discussions and contributions to these images. The rocker project wiki
provides a place to synthesize community-contributed documentation,
use-cases, and other knowledge about using the Rocker images.

@@

All Rocker images are based on the Debian Linux distribution.
The Debian platform provides a small base image, the well-known `apt`
package management system and rich ecosystem of software libraries,
making it the base image of choice for Docker images, including many
of the "official" images maintained by Docker's own development team.
The Debian platform is also perhaps the best supported Linux platform
within the R community, including an active `r-sig-debian` listserve.
The relatively long period between stable Debian releases
(roughly two years recently) means that software in the Debian stable 
(e.g. `debian:jessie`, `debian:stretch`)
releases can lag significantly behind current releases of popular
software, including R.  More recent versions of packages can be found in
the pre-release distribution, `debian:testing`, while the very latest
binary builds can be found on `debian:unstable`.   The Rocker project
can be largely divided into two stacks which address different needs,
reflected in which Debian distribution they are based on.  The first stack
is based on `debian:testing`, the second, more recently introduced stack
is based only on Debian stable releases.  Rocker images always point to specific
stable releases (`jessie`, `stretch`), and do not use the tag `debian:stable`,
which is a rolling tag that always points to the most
recent stable version.   The different Rocker stacks have different
aims and thus provide different images, as shown in Tables 1 & 2.

@@

As these images focus on developers and/or as base images for custom uses,
this stack does not include many specific R packages. Additional dependencies
and packages can easily be installed from `apt`.  R packages not available
in the `apt` repositories can be installed directly from CRAN using the
`littler` scripts, as described in <https://rocker-project.org/use>.

This stack also include the images `shiny` and `rstudio:testing` that
provide Shiny server and RStudio\textsuperscript{\textregistered} server IDE from RStudio\textsuperscript{\textregistered} Inc, built on the
`r-base` image.  RStudio\textsuperscript{\textregistered} and Shiny are registered trademarks of RStudio
Inc, and their use and the distribution of their software in binary form on
Docker Hub has been granted to the Rocker project by explicit permission
from RStudio. Users should review RStudio's trademark use policy
(<http://www.rstudio.com/about/trademark/>) and address inquiries about
further distribution or other questions to <permissions@rstudio.com>.
The Rocker project also provides images with RStudio\textsuperscript{\textregistered} server and Shiny
server in the stable versioned stack.

@@

@@

**Installation**: In this stack, the desired version of R is always built
directly from source rather than the `apt` repositories.  Compilers and
dependencies are still installed from the stable `apt` repositories,
and thus lag behind the more recent versions found in the `testing`
stack. Version tags `3.3.3` and older are based on the Debian 8.0 release,
code-named `jessie`, while `3.4.0` - `3.4.2`, `devel`, and `latest` are based on
Debian 9.0, `stretch`, (released 2017-06-17, while R was at `3.4.0`), 
and thus have access to much
newer versions of common system dependencies and compilers.  Dependencies
needed to compile R that are not required at runtime are removed once R is
installed, keeping the base images light-weight for faster download times.
While most system dependencies required by common R packages can still
be installed from the `apt` repositories, occasionally a more recent
version must be compiled from source (e.g. the Gibbs Sampling library,
JAGS, and the geospatial toolkit GDAL, must both be compiled from source
on `debian:jessie` images).  In this stack, users should avoid installing R
packages using `apt` without careful consideration, as this will install a second, 
probably different version
of R from the Debian repositories and a dated version of the R package
(since any `r-cran-pkgname` package in the Debian repositories will
depend on `r-base` in `apt` as well).

@@

Over the past several years, Docker has seen immense adoption across industry and academia. 
The Open Container initiative \citep{oci} now provides an open standard that has further
extended this container approach to research environments through projects such as 
Singularity \citep{singularity}, allowing users to deploy containerized environments such as Rocker on machines where they do not have root access, such as clusters or private servers. Containerization promises
to solve numerous challenges such as portability and replicability in research computing,
which often relies on complex and heterogeneous software stacks \citep{Boettiger2015}. Yet
implementing such environments in containers is not a trivial task, and not all implementations
provide the same usability, portability or reproducibility.  Here we have detailed the 
approach taken by the Rocker project in creating and maintaining these environments through
an open and community-driven process.  This structure of the Rocker project has evolved
over three years of operation while drawing in an ever widening base of academic researchers, 
university instructors and industry users.  We believe this overview will be instructive
not only to users and developers interested in the Rocker project, but as a model for
similar efforts around other environments or domains.

>>>>

Docker is a popular open-source tool to create, distribute, deploy,
and run software applications using *containers*.  Containers provide a
virtual environment (see \citet{Clark2014} for an overview of common virtual environments) requiring all operating-system components an application needs to run:
code, runtime, system tools, system runtime.  Docker containers are
lightweight as they share the operating system kernel, starting instantly
using a layered filesystem which minimizes disk footprint and download time,
are built on open standards that run on all major platforms (Linux, Mac,
Windows), and provide an added layer of security by running an application
in an isolated environment \citep{what-docker}.  Familiarity with a few key
terms is helpful in understanding this paper. The term "container" refers to 
an isolated software environment on a computer. R users can think of running
a container as analogous to loading an R package; it makes a set of software
functions available.  A Docker "image" is a binary archive of that software,
analogous to an R binary package -- a given version need be downloaded only
once and then can be "run" to create a container whenever it is needed.
A "Dockerfile" is a recipe, the source-code, to create a Docker image.  Development
and contributions to the Rocker project focuses on the construction, organization
and maintenance of these Dockerfiles.

@@

<!-- DE: It is a little weird to have this first as it is a somewhat
advanced, and I'd conjecture "proportionally speaking, "rare" use.
Isn't the main use to just "deploy an application" _locally_? I how we
hand users R, or R-devel, or RStudio, or Tibbliesverse ?

By the same token, it is a bit weird to intro Singularity _here_ and I
still don't know if Shifter really needs to be in the paper (cf
Patrick's comments)

So should we maybe insert a really short para like this

   Deployment: Ready-to-run binaries

   Docker permits convenient access to pre-configured and create
   _binary_ images that "just work".  This allows R users to access a
   wider-variety of ready-to-use environments than the R Project
   itself provides.  For example, R users on Windows may run RStudio
   Server or Shiny Server locally just by launching a single command
   (once Docker itself is installed), access R-devel without affecting
   their local system and more.

nd then start the para with 'Another' instead on one?  There is a
little overlap with 'Sandboxed' below but I think it is worth it to
start with "first things first" here

@@

Importantly, Rocker containers are just as easy to run locally on most modern laptops
by first installing the appropriate Docker command (or distribution) for Windows, MacOS, and Linux-based
operating systems. By sharing volumes with the local host, users can still manipulate
files with familiar, native tools while performing computation through a reproducible,
containerized environment \citep{Boettiger2015}. Being
able to test code in a predictable, pre-configured R environment on a
local machine and to then run the same code in an identical environment on a
remote server (e.g. for access to greater RAM, more processors, or merely
to free up the local machine from a long-running computation) is essential
for low-friction scaling of analysis.  Without such containerization,
getting code to run appropriately in a remote environment can be a major
undertaking, requiring both time and knowledge many would-be users may
not have.

@@

This sandboxing feature is also valuable in the 
remote computing context, allowing system administrators to grant
users freedom to install software which requires root privileges inside a container,
while not granting them root access on the host machine. Root access
is required to launch Docker containers, though not to access containers already
<!-- DE False. On Linux the dockerd has the correct rights, and I (as member of group Docker) 
can access docker without it.  That is important: minimal sudo use!  -->
running and providing some service such as RStudio. Users logging into a container
through the RStudio\textsuperscript{\textregistered} interface do not by default have root privileges, though are able
to install R packages. Granting these users root privileges in the container still leaves
them sandboxed from the host container.  Further, Docker images can
be deployed from a machine without root privileges through the alternate container
runtime environment, Singularity \citep{singularity}.  Unlike traditional virtual machines,
these containers do not impose a large footprint of reserved resources --
a typical host can easily support 100s of containers \citep{what-docker}.

@@

Users building on the version-tagged images will by default use the MRAN snapshot mirror \citep{MRAN}
associated with the most recent date for which that image was current.  This ensures that
a Dockerfile building `FROM rocker/verse:3.4.0` will only install R package versions that were
available on CRAN on 2017-06-30, that is, the day R v3.4.1 was released.  This default
can of course be overwritten in the standard R manner, e.g. by specifying a different CRAN mirror explicitly in any 
command to `install.packages()`, or by adjusting the default CRAN mirror in `options(repo=<CRAN-MIRROR>)`
in an `.Rprofile`.  Note that the MRAN date associated with the current release (e.g. `3.4.1` at the time of
writing) will continue to advance on the Docker-hub image until the next R release.  Software
installed from `apt-get` in these images will come from the the stable Debian release (`stretch` or `jessie`)
and thus not change versions (though it will receive security patches).  Packages installed from BioConductor
using `bioclite()` utility will also install the version appropriate to the version of R found on the system
(the Bioconductor semi-annual release model avoids the need for an MRAN mirror). Users installing packages
from GitHub or other sources can request a specific git release tag or hash for a more reproducible build,
or adopt an alternative approach such as `packrat` \citep{packrat}. A more general discussion of the use and limitations of Docker for computational reproducibility can be found in \cite{Boettiger2015}. 
<!--
however, `packrat` should not be necessary
for just maintaining reproducibility with respect to CRAN and Bioconductor stacks installed on versioned images.
-->

@@

Any portable computational environment faces an inevitable tension between
the "kitchen sink problem" at one extreme, and the "discovery problem" on
the other.  A kitchen sink image seeks to accommodate too many use cases
in a single image.  Such images are inevitably very large and thus slow
or difficult to deploy, maintain and optimize.  At the other extreme, providing
too many specialized images makes it more difficult for a user to discover 
the one they need.  The Rocker project seeks to avoid both of these problems
by providing a carefully-curated suite of images that an be easily extended by
individuals and communities.

@@

The Rocker project consists of a suite of images built automatically by
and hosted on the Docker Hub, <https://hub.docker.com/r/rocker>.  Source
Dockerfiles, supporting scripts and documentation are hosted on GitHub
under the organization `rocker-org`, <https://github.com/rocker-org>.
The issue tracker and pull requests are used for community input,
discussions and contributions to these images. The Rocker project wiki
provides a place to synthesize community-contributed documentation,
use-cases, and other knowledge about using the Rocker images.

@@

All Rocker images are based on the Debian Linux distribution.
The Debian platform provides a small base image, the well-known `apt`
package management system and rich ecosystem of software libraries,
making it the base image of choice for Docker images, including many
of the "official" images maintained by Docker's own development team.
The Debian platform is also perhaps the best supported Linux platform
within the R community, including an active `r-sig-debian` listserve.
The relatively long period between stable Debian releases
(roughly two years recently) means that software in the Debian stable 
(e.g. `debian:jessie`, `debian:stretch`)
releases can lag significantly behind current releases of popular
software, including R.  More recent versions of packages can be found in
the pre-release distribution, `debian:testing`, while the very latest
binary builds can be found on `debian:unstable`.   The Rocker project
can be largely divided into two stacks which address different needs,
reflected in which Debian distribution they are based on.  The first stack
is based on `debian:testing`. The second, more recently introduced stack
is based only on Debian stable releases.  Rocker images always point to specific
stable releases (`jessie`, `stretch`), and do not use the tag `debian:stable`,
which is a rolling tag that always points to the most
recent stable version.   The different Rocker stacks have different
aims and thus provide different images, as shown in Tables 1 & 2 below.

@@

As these images focus on developers and/or as base images for custom uses,
this stack does not include many specific R packages. Additional dependencies
and packages can easily be installed from `apt`.  R packages not available
in the `apt` repositories can be installed directly from CRAN using either `R` or 
the `littler` scripts, as described in <https://rocker-project.org/use>.

This stack also includes the images `shiny` and `rstudio:testing` that
provide Shiny server and RStudio\textsuperscript{\textregistered} server IDE from RStudio\textsuperscript{\textregistered} Inc, built on the
`r-base` image.  RStudio\textsuperscript{\textregistered} and Shiny are registered trademarks of RStudio
Inc, and their use and the distribution of their software in binary form on
Docker Hub has been granted to the Rocker project by explicit permission
from RStudio. Users should review RStudio's trademark use policy
(<http://www.rstudio.com/about/trademark/>) and address inquiries about
further distribution or other questions to <permissions@rstudio.com>.
The Rocker project also provides images with RStudio\textsuperscript{\textregistered} server and Shiny
server in the stable versioned stack.

@@

<!-- DE The (+46,000 as hadleyverse) is visually very awkward. Can the "+46k" be a footnote? I know nothing about Markdown tables but would be happy to redo 1 + 2 in latex ...-->

@@

**Installation**: In this stack, the desired version of R is always built
directly from source rather than the `apt` repositories.  Compilers and
dependencies are still installed from the stable `apt` repositories,
and thus lag behind the more recent versions found in the `testing`
stack. Version tags `3.3.3` and older are based on the Debian 8.0 release,
code-named `jessie`, while `3.4.0` - `3.4.2`, `devel`, and `latest` are based on
Debian 9.0, `stretch`, (released 2017-06-17, while R was at `3.4.0`), 
and thus have access to much
newer versions of common system dependencies and compilers.  Dependencies
needed to compile R that are not required at runtime are removed once R is
installed, keeping the base images light-weight for faster download times.
While most system dependencies required by common R packages can still
be installed from the `apt` repositories, occasionally a more recent
version must be compiled from source (e.g. the Gibbs Sampling program
JAGS \citep{jags}, and the geospatial toolkit GDAL, must both be compiled from source
on `debian:jessie` images).  In this stack, users should avoid installing R
packages using `apt` without careful consideration, as this will install a second, 
probably different version
of R from the Debian repositories and a dated version of the R package
(since any `r-cran-pkgname` package in the Debian repositories will
depend on `r-base` in `apt` as well).

@@

Over the past several years, Docker has seen immense adoption across industry and academia. 
The Open Container initiative \citep{oci} now provides an open standard that has further
extended this container approach to research environments through projects such as 
Singularity \citep{singularity}, allowing users to deploy containerized environments such as Rocker on machines where they do not have root access, such as clusters or private servers. Containerization promises
to solve numerous challenges such as portability and replicability in research computing,
which often relies on complex and heterogeneous software stacks \citep{Boettiger2015}. Yet
implementing such environments in containers is not a trivial task, and not all implementations
provide the same usability, portability or reproducibility.  Here we have detailed the 
approach taken by the Rocker project in creating and maintaining these environments through
an open and community-driven process.  This structure of the Rocker project has evolved
over three years of operation while drawing in an ever-widening base of academic researchers, 
university instructors and industry users.  We believe this overview will be instructive
not only to users and developers interested in the Rocker project, but as a model for
similar efforts around other environments or domains.
